Assimilating disturbance: Toward real-time monitoring and forecasting

### 2023-02-01 results.Rmd
formatting

```{r sssd, echo=FALSE,fig.cap="Experiment 2: Assimilation using process-based VSEM captures multi-pool disturbance to leaf and wood biomass"}

@@

load("exp2.Rdata") # Generated by SingleSite_SingleDisturb_MultiPool.Rmd
#Figures showing all 3 pools cov, C accounting / attribution 
pool.names=c("Cleaf","Csoil","Cwood")
for(i in 1:3){

  CI.Na = sapply(X = Analysis,function(y){
    x = y$MCMC
    sel.n = which(colnames(x) == paste0("n[",i,"]"))
    quantile(x[,sel.n],c(0.025,0.5,0.975))
  })
  D.Na = sapply(X = Analysis,function(y){
    x = y$MCMC
    sel.n = which(colnames(x) == paste0("D"))
    mean(x[,sel.n])
  })

  rng = range(CI.Na)
  if(rng[1]<1) rng[1] = 0
  plot(No[,i],
       ylim = rng,type='n',
       ylab=pool.names[i],xlab="time")
  ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[3,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
  lines(CI.Na[2,],col="#b2df8a",lwd=2)
  points(No[,i])
  #lines(Ntrue[,i],lty=2,lwd=2)
  #lines(D.Na,lwd=2)
  if(i==1){
    legend("bottomright",legend=c("Assimilation","pseudodata"),
       lwd=5,col=c("#b2df8a","black"))
  }
}
```

@@

plot(Analysis[[10]]$Na[,1],Analysis[[10]]$Na[,3],pch="+",col=ecoforecastR::col.alpha("#b2df8a",alpha=1),
     ylim=range(Forecast[[10]]$Nf[,3]),
     xlim=range(Forecast[[10]]$Nf[,1]),
     xlab=pool.names[1],ylab=pool.names[3],
     cex.lab=1.25,asp = 1)
points(Forecast[[10]]$Nf[,1],Forecast[[10]]$Nf[,3],cex=1,col="purple",pch=18)
points(No[10,1],No[10,3],cex=3,col=1)
lines(No[10,1]+c(-1,1)*sqrt(R[1,1]),rep(No[10,3],2))
lines(rep(No[10,1],2),No[10,3]+c(-1,1)*sqrt(R[3,3]))

@@

```{r multiPseudo, echo=FALSE, message = FALSE, fig.cap="Simulated experiments 3-5: Left hand column shows timeseries for the wood pool for all three experiments; Right hand shows bivariate ensemble adjustment at time of disturbance",fig.height=4,fig.width=3}

@@

########experiment 3 - two disturb
load("exp3.Rdata") ## Generated by SingleSite_MultiDisturb_MultiPool.Rmd
pool.names=c("Cleaf","Csoil","Cwood")
i = 3
CI.Na = sapply(X = Analysis,function(y){
  x = y$MCMC
  sel.n = which(colnames(x) == paste0("X[",i,"]"))
  quantile(x[,sel.n],c(0.025,0.5,0.975))
})
rng = range(CI.Na)
if(rng[1]<1) rng[1] = 0
plot(No[,i],
     ylim = rng,type='n',
     ylab=pool.names[i],xlab="time",
     main="Multi-Disturbance")
ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[3,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
lines(CI.Na[2,],col="#b2df8a",lwd=2)
points(No[,i])
legend("bottomleft",legend=c("Assimilation","pseudodata"),
       lwd=5,col=c("#b2df8a","black"))

plot(Analysis[[10]]$Na[,1],Analysis[[10]]$Na[,3],pch=Analysis[[10]]$Ca,col=ecoforecastR::col.alpha("#b2df8a",alpha=1),
     ylim=range(Forecast[[10]]$Nf[,3]),
     xlim=range(Forecast[[10]]$Nf[,1]),
     xlab=pool.names[1],ylab=pool.names[3],
     cex.lab=1.25,asp = 1)
points(Forecast[[10]]$Nf[,1],Forecast[[10]]$Nf[,3],cex=1,col="purple",pch=18)
points(No[10,1],No[10,3],cex=3,col=1)
lines(No[10,1]+c(-1,1)*sqrt(R[1,1]),rep(No[10,3],2))
lines(rep(No[10,1],2),No[10,3]+c(-1,1)*sqrt(R[3,3]))

@@

##### experiment 4 - two disturb, pft switch
load("exp4.Rdata") ## SingleSite_MultiDisturb_MultiPool_MultiPFT.Rmd
pool.names=c("Cleaf","Csoil","Cwood")
i = 3
CI.Na = sapply(X = Analysis,function(y){
  x = y$MCMC
  sel.n = which(colnames(x) == paste0("X[",i,"]"))
  quantile(x[,sel.n],c(0.025,0.5,0.975))
})
rng = range(CI.Na)
if(rng[1]<1) rng[1] = 0
plot(No[,i],
     ylim = rng,type='n',
     ylab=pool.names[i],xlab="time",
     main="Multi-Disturbance, PFT switching")
ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[3,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
lines(CI.Na[2,],col="#b2df8a",lwd=2)
points(No[,i])
legend("bottomleft",legend=c("Assimilation","pseudodata"),
       lwd=5,col=c("#b2df8a","black"))

plot(Analysis[[10]]$Na[,1],Analysis[[10]]$Na[,3],pch=Analysis[[10]]$Ca,col=ecoforecastR::col.alpha("#b2df8a",alpha=1),
     ylim=range(Forecast[[10]]$Nf[,3]),
     xlim=range(Forecast[[10]]$Nf[,1]),
     xlab=pool.names[1],ylab=pool.names[3],
     cex.lab=1.25)
points(Forecast[[10]]$Nf[,1],Forecast[[10]]$Nf[,3],cex=1,col="purple",pch=18)
points(No[10,1],No[10,3],cex=3,col=1)
lines(No[10,1]+c(-1,1)*sqrt(R[1,1]),rep(No[10,3],2))
lines(rep(No[10,1],2),No[10,3]+c(-1,1)*sqrt(R[3,3]))

@@

######### experiment 5 - one disturb, 2 sites
load("exp5trim.Rdata") ##Generated by TwoSite_SingleDisturb_MultiPool.Rmd
pool.names=c("Cleaf","Csoil","Cwood")
rng = range(CI.Na)
if(rng[1]<1) rng[1] = 0
plot(No[,i],
     ylim = rng,type='n',
     ylab=pool.names[i],xlab="time",
     main="Multi-Site, Single-Disturbance")
  ecoforecastR::ciEnvelope(1:ncol(CI.Na[[3]]),CI.Na[[3]][1,],CI.Na[[3]][3,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
  ecoforecastR::ciEnvelope(1:ncol(CI.Na[[6]]),CI.Na[[6]][1,],CI.Na[[6]][3,],col=ecoforecastR::col.alpha("green",0.7))
lines(CI.Na[[3]][2,],col="#b2df8a",lwd=2)
lines(CI.Na[[6]][2,],col="green",lwd=2)
points(No[,3])
points(No[,6],pch=2)
legend("bottomleft",legend=c("Site 1","Site 2","pseudodata"),
       lwd=5,col=c("#b2df8a","green","black"))

### cov plot
ss = floor(seq(1,nrow(Analysis10$Na),length=200)) ## subsample
plot(Analysis10$Na[,1],Analysis10$Na[,3],pch=Analysis10$Ca,col=ecoforecastR::col.alpha("#b2df8a",alpha=1),
     ylim=range(Forecast10$Nf[,3,1]),
     xlim=range(Forecast10$Nf[,1,1]),
     xlab=pool.names[1],ylab=pool.names[3],
     cex.lab=1.25,asp = 1,type='n')
points(Forecast10$Nf[ss,1,1],Forecast10$Nf[ss,3,1],cex=0.5,col="purple",pch=18)
points(Forecast10$Nf[ss,1,2],Forecast10$Nf[ss,3,2],cex=0.5,col="violet",pch=18)
points(Analysis10$Na[ss,1],Analysis10$Na[ss,3],pch=Analysis10$Ca[ss],col=ecoforecastR::col.alpha("#b2df8a",alpha=0.8),cex=0.5)
points(Analysis10$Na[ss,4],Analysis10$Na[ss,6],pch=Analysis10$Ca[ss],col=ecoforecastR::col.alpha("green",alpha=0.8),cex=0.5)
points(No[10,1],No[10,3],cex=3,col=1)
lines(No[10,1]+c(-1,1)*sqrt(R[1,1]),rep(No[10,3],2))
lines(rep(No[10,1],2),No[10,3]+c(-1,1)*sqrt(R[3,3]))
points(No[10,4],No[10,6],cex=3,col=1)
lines(No[10,4]+c(-1,1)*sqrt(R[1,1]),rep(No[10,6],2))
lines(rep(No[10,4],2),No[10,6]+c(-1,1)*sqrt(R[3,3]))

@@

Prior to applying the VSEM to simulate the conifer forest carbon cycle in Oregon the model was calibrated against AGB, leaf C, and NEE data from Ameriflux site US-ME2.  Best fit parameters are given in Supplemental Table \@ref(tab:parameters). The best fit to the observed data was for AGB, which had a $R^2$ of 0.977, a RMSE of 0.089 $kgC/m^2$ (1.7%), and a bias of -1.1% (-0.059 $kgC/m^2$) (Supplemental figure \@ref(fig:calibrate) ). Achieving this fit came at the cost of nontrivial systematic biases in both leaf carbon (0.064 $kgC/m^2$, 13%) and NEE (0.528 $kgC/m^2/year$, 69%). The RMSEs for leaf carbon (0.067 $kgC/m^2$, 13.6%) and NEE (0.612 $kgC/m^2/year$, 80%) were similar to the biases, suggesting that most of the model error was systematic, particularly for leaf carbon.

@@

```{r ORorig, echo=FALSE, fig.asp=0.75, fig.cap="Filter divergence and disturbance detection for central Oregon test case. Left column is the default configuration while the right column is a reduced observation error experiment (Exp 2). Top row is the assimilation for an example site. Middle row shows the across-site distribution of divergence metrics and how these vary by disturbance magnitude (purple) and disturbance type (symbols). Bottom row shows how the probability of detecting the disturbance varies with disturbance magnitude"}
## summary figure showing didn't work great at default error rate
m = matrix(1:6,nrow=3,byrow = TRUE)
layout(m)

### Row 1: Example time-series ###
site = 188
exp.name = c("P1Q1R1","P1Q1R4")
exp.pretty.name = c("(a) Example site, Original","(c) Example site, Low Obs Error")
pool.names=c("Cleaf","Csoil","AGB")
i = 3 ## Stem
time = 1985:2017
for(ex in seq_along(exp.name)){
  load(paste0("OR_TS_",site,"_",exp.name[ex],".RData"))
  plot(time,Yi[,i],
       ylim = range(rbind(CI.Na,Yi[,i],CI.D),na.rm=TRUE),
       type='n',
       ylab=pool.names[i],xlab="time",
       main=exp.pretty.name[ex])
  ecoforecastR::ciEnvelope(time,CI.Na[1,],CI.Na[5,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
  lines(time,CI.Na[3,],col="#b2df8a",lwd=2)
  points(time,Yi[,i])
  if(ex == 1) {
    legend("bottomleft",legend=c("Assimilation","Data"),
       lwd=5,col=c("#b2df8a","black"),cex=0.7)
  }
}

@@

### Row 3: Disturbance magnitude vs P(detection) ###
pred.seq=list()
distabs = obs.pre-obs.dist
pred.seq$distabs = seq(min(distabs),max(distabs),length=500)
detect.pretty.name = c("(c) Probability of detection, Original","(f) Probability of Detection, Low Obs Error")
dist.qb.abs <- list()
for(i in 1:2){
  distrate = distrate = elist[[i]]$distrate
  dist.qb.abs[[i]] = glm(distrate ~ distabs,
                    na.action=na.omit,
                    family=quasibinomial("logit")
  )
  plot(distabs,distrate,col=disturb.col.al[obs.class.re],
       xlab="AGB disturbance magnitude",ylab="P(detect)",
       pch=16, main=detect.pretty.name[i],xlim=c(0,17))
  lines(pred.seq$distabs,predict(dist.qb.abs[[i]],newdata=pred.seq,type="response"),col="purple")
  if(i == 1) legend("topleft",legend=disturb.names,
                  col=disturb.col,pch=16,cex=0.7,
                  text.col=c(1,disturb.col[2:5]))
}
```
In total we assimilated 365 example AGB timeseries from our central Oregon study region, with 81 coming from undisturbed sites and 78, 88, 92, and 17 coming from clear cuts, fires, forest pests, and other disturbances, respectively. Initial applications of the multinomial filter exhibited high rates of filter divergence, with a coverage of only `r format(stats[[1]]$div.cov*100,digits=3)`% of the observations falling within the Analysis 95% CI at the end of the timeseries. Filter divergence is a known issue for Kalman Filters, not something unique to our multinomial filter, and results from the model becoming falsely overconfident in itself such that it ignores, and diverges from, the data. Figure \@ref(fig:ORorig) a demonstates an example of this for a site that experienced a forest fire in 1999, yet the model continues to believe the site is undisturbed. Across sites the magnitude of filter divergence observed was negatively correlated with the amount of biomass at the end of the timeseries (Figure \@ref(fig:ORorig) b, p < `r format(anova(div.reg[[1]])$"Pr(>F)"[1],digits=3,scientific=TRUE)`). In other words, sites with low final biomass (i.e., large previous disturbance) showed a larger divergence and the divergence declined as the final biomass increased. Divergence also varied by disturbance type (p < `r format(anova(div.class[[1]])$"Pr(>F)"[2],digits=3,scientific=TRUE)`), with clear cuts having the lowest divergence for a given disturbance size and forest pests being lower than the fires and "other". Despite the issue of filter divergence there was still a positive relationship between disturbance magnitude and the posterior probability of detecting disturbance (Figure \@ref(fig:ORorig) c, p < `r format(summary(dist.qb.abs[[1]])$coefficients[2,4],digits=3,scientific=TRUE)`). However, detection was rarely successful for AGB disturbances that removed less than 7.5 $kgC/m^2$ and did not exceed 50% until disturbances were larger than `r bm=coef(dist.qb.abs[[1]]); format(-bm[1]/bm[2],digits=2)` $kgC/m^2$.

The causes of filter divergence are known to be some combination of model overconfidence (e.g., insufficient process error, Q) and high observation error (R) in the data. As described in the Methods we ran additional computational experiments where we varied Q and R (\@ref(tab:QRexp)) to verify that without filter divergence the multinomial filter was able to capture real-world disturbances and to assess what level of variance inflation was required to do so. Figure \@ref(fig:ORorig) d-f show the results for Experiment 2, where we experimented with reducing the observation error, R. In this experiment we see that the data assimilation follows the observations closely, with a much tighter interval estimate, and is able to catch the disturbance event (Figure \@ref(fig:ORorig) d). We also see much lower incidence of filter divergence across sites, with a posterior coverage at the final timestep of `r format(stats[[2]]$div.cov*100,digits=3)`%, (Figure \@ref(fig:ORorig) e). Divergence is no longer significantly affected by disturbance magnitude (p < `r format(summary(div.class[[2]])$coefficients[2,4],digits=3)`) and the only significant effect of disturbance type was a small reduction in divergence for clearcuts (p < `r format(summary(div.class[[2]])$coefficients[3,4],digits=3)`). Most importantly, detection probabilities were strongly related to disturbance magnitude (Figure \@ref(fig:ORorig) f), successfully detecting disturbances that removed more than `r bm=coef(dist.qb.abs[[2]]); format(-bm[1]/bm[2],digits=2)` $kgC/m^2$.

@@

```{r divExp,echo=FALSE,fig.cap="Sensitivity of (a) filter divergence coverage, and (b) disturbance detection threshold to variance inflation of the observation error (x-axis) and process error (symbols)",fig.asp=0.75 }
m = matrix(1:3,nrow=3,byrow = TRUE)
layout(m)
## Divergence
div.cov = sapply(stats,"[[","div.cov")
plot(Rmult[Rs],div.cov,col=Qs,pch=Qs,log="x",xlab="Variance Multiplier",ylab="Filter Coverage",main="(a) Filter Coverage")
legend("topright",legend=paste0(Qmult[1:2],"x"),col=1:2,pch=1:2,title="Q")

@@

## Detection probability
for(i in 3:6){
  distrate = distrate = elist[[i]]$distrate
  dist.qb.abs[[i]] = glm(distrate ~ distabs,
                    na.action=na.omit,
                    family=quasibinomial("logit")
  )
}
thresh = sapply(dist.qb.abs,function(x){bm=coef(x); -bm[1]/bm[2]})
plot(Rmult[Rs],thresh,col=Qs,pch=Qs,log="x",xlab="Variance Multiplier",ylab=expression("Detection Threshold" ~ (kgC/m^2)),main="(c) Disturbance Detection Threshold")
legend("topleft",legend=paste0(Qmult[1:2],"x"),col=1:2,pch=1:2)
```

>>>>

```{r sssd, echo=FALSE,fig.asp=1.25,fig.cap="Experiment 2: Assimilation using process-based VSEM captures multi-pool disturbance to leaf and wood biomass"}

@@

load("exp2.Rdata") # Generated by SingleSite_SingleDisturb_MultiPool.Rmd
#Figures showing all 3 pools cov, C accounting / attribution 
pool.names=c("Leaf","Soil","Wood")
scaled = c(1,10,1)
for(i in 1:3){

  CI.Na = sapply(X = Analysis,function(y){
    x = y$MCMC
    sel.n = which(colnames(x) == paste0("n[",i,"]"))
    quantile(x[,sel.n],c(0.025,0.5,0.975))
  })
  D.Na = sapply(X = Analysis,function(y){
    x = y$MCMC
    sel.n = which(colnames(x) == paste0("D"))
    mean(x[,sel.n])
  })
  CI.Na = CI.Na/scaled[i]

  rng = range(CI.Na)
  if(rng[1]<1) rng[1] = 0
  plot(No[,i]/scaled[i],
       ylim = rng,type='n',
       ylab=bquote(.(pool.names[i]) ~ (kgC/m^2)),xlab="time")
  ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[3,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
  lines(CI.Na[2,],col="#b2df8a",lwd=2)
  points(No[,i]/scaled[i])
  #lines(Ntrue[,i],lty=2,lwd=2)
  #lines(D.Na,lwd=2)
  if(i==1){
    legend("bottomright",legend=c("Assimilation","pseudodata"),
       lwd=5,col=c("#b2df8a","black"))
  }
}
```

@@

plot(Analysis[[10]]$Na[,1],Analysis[[10]]$Na[,3],pch="+",col=ecoforecastR::col.alpha("#b2df8a",alpha=1),
     ylim=range(Forecast[[10]]$Nf[,3]),
     xlim=range(Forecast[[10]]$Nf[,1]),
     xlab=bquote(.(pool.names[1]) ~ (kgC/m^2)),ylab=bquote(.(pool.names[3])~(kgC/m^2)),
     cex.lab=1.25,asp = 1)
.(labNames[1]) ~ x^2
points(Forecast[[10]]$Nf[,1],Forecast[[10]]$Nf[,3],cex=1,col="purple",pch=18)
points(No[10,1],No[10,3],cex=3,col=1)
lines(No[10,1]+c(-1,1)*sqrt(R[1,1]),rep(No[10,3],2))
lines(rep(No[10,1],2),No[10,3]+c(-1,1)*sqrt(R[3,3]))

@@

```{r multiPseudo, echo=FALSE, message = FALSE, fig.cap="Simulated experiments 3-5: Left hand column shows timeseries for the wood pool for all three experiments; Right hand shows bivariate ensemble adjustment at time of disturbance",fig.asp=1.25}

@@

########experiment 3 - two disturb
load("exp3.Rdata") ## Generated by SingleSite_MultiDisturb_MultiPool.Rmd
pool.names=c("Cleaf","Csoil","Cwood")
i = 3
CI.Na = sapply(X = Analysis,function(y){
  x = y$MCMC
  sel.n = which(colnames(x) == paste0("X[",i,"]"))
  quantile(x[,sel.n],c(0.025,0.5,0.975))
})
rng = range(CI.Na)
if(rng[1]<1) rng[1] = 0
plot(No[,i],
     ylim = rng,type='n',
     ylab=bquote(.(pool.names[i]) ~ (kgC/m^2)),xlab="time",
     main="Multi-Disturbance")
ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[3,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
lines(CI.Na[2,],col="#b2df8a",lwd=2)
points(No[,i])
legend("bottomleft",legend=c("Assimilation","pseudodata"),
       lwd=5,col=c("#b2df8a","black"))

plot(Analysis[[10]]$Na[,1],Analysis[[10]]$Na[,3],pch=Analysis[[10]]$Ca,col=ecoforecastR::col.alpha("#b2df8a",alpha=1),
     ylim=range(Forecast[[10]]$Nf[,3]),
     xlim=range(Forecast[[10]]$Nf[,1]),
     xlab=bquote(.(pool.names[1]) ~ (kgC/m^2)),ylab=bquote(.(pool.names[3]) ~ (kgC/m^2)),
     cex.lab=1.25,asp = 1)
points(Forecast[[10]]$Nf[,1],Forecast[[10]]$Nf[,3],cex=1,col="purple",pch=18)
points(No[10,1],No[10,3],cex=3,col=1)
lines(No[10,1]+c(-1,1)*sqrt(R[1,1]),rep(No[10,3],2))
lines(rep(No[10,1],2),No[10,3]+c(-1,1)*sqrt(R[3,3]))

@@

##### experiment 4 - two disturb, pft switch
load("exp4.Rdata") ## SingleSite_MultiDisturb_MultiPool_MultiPFT.Rmd
pool.names=c("Cleaf","Csoil","Cwood")
i = 3
CI.Na = sapply(X = Analysis,function(y){
  x = y$MCMC
  sel.n = which(colnames(x) == paste0("X[",i,"]"))
  quantile(x[,sel.n],c(0.025,0.5,0.975))
})
rng = range(CI.Na)
if(rng[1]<1) rng[1] = 0
plot(No[,i],
     ylim = rng,type='n',
     ylab=bquote(.(pool.names[i]) ~ (kgC/m^2)),xlab="time",
     main="Multi-Disturbance, PFT switching")
ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[3,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
lines(CI.Na[2,],col="#b2df8a",lwd=2)
points(No[,i])
legend("bottomleft",legend=c("Assimilation","pseudodata"),
       lwd=5,col=c("#b2df8a","black"))

plot(Analysis[[10]]$Na[,1],Analysis[[10]]$Na[,3],pch=Analysis[[10]]$Ca,col=ecoforecastR::col.alpha("#b2df8a",alpha=1),
     ylim=range(Forecast[[10]]$Nf[,3]),
     xlim=range(Forecast[[10]]$Nf[,1]),
     xlab=bquote(.(pool.names[1]) ~ (kgC/m^2)),
     ylab=bquote(.(pool.names[3]) ~ (kgC/m^2)),
     cex.lab=1.25)
points(Forecast[[10]]$Nf[,1],Forecast[[10]]$Nf[,3],cex=1,col="purple",pch=18)
points(No[10,1],No[10,3],cex=3,col=1)
lines(No[10,1]+c(-1,1)*sqrt(R[1,1]),rep(No[10,3],2))
lines(rep(No[10,1],2),No[10,3]+c(-1,1)*sqrt(R[3,3]))

@@

######### experiment 5 - one disturb, 2 sites
load("exp5trim.Rdata") ##Generated by TwoSite_SingleDisturb_MultiPool.Rmd
pool.names=c("Cleaf","Csoil","Cwood")
rng = range(CI.Na)
if(rng[1]<1) rng[1] = 0
plot(No[,i],
     ylim = rng,type='n',
     ylab=bquote(.(pool.names[i]) ~ (kgC/m^2)),xlab="time",
     main="Multi-Site, Single-Disturbance")
  ecoforecastR::ciEnvelope(1:ncol(CI.Na[[3]]),CI.Na[[3]][1,],CI.Na[[3]][3,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
  ecoforecastR::ciEnvelope(1:ncol(CI.Na[[6]]),CI.Na[[6]][1,],CI.Na[[6]][3,],col=ecoforecastR::col.alpha("green",0.7))
lines(CI.Na[[3]][2,],col="#b2df8a",lwd=2)
lines(CI.Na[[6]][2,],col="green",lwd=2)
points(No[,3])
points(No[,6],pch=2)
legend("bottomleft",legend=c("Site 1","Site 2","pseudodata"),
       lwd=5,col=c("#b2df8a","green","black"))

### cov plot
ss = floor(seq(1,nrow(Analysis10$Na),length=200)) ## subsample
plot(Analysis10$Na[,1],Analysis10$Na[,3],pch=Analysis10$Ca,col=ecoforecastR::col.alpha("#b2df8a",alpha=1),
     ylim=range(Forecast10$Nf[,3,1]),
     xlim=range(Forecast10$Nf[,1,1]),
     xlab=bquote(.(pool.names[1]) ~ (kgC/m^2)),
     ylab=bquote(.(pool.names[3]) ~ (kgC/m^2)),
     cex.lab=1.25,asp = 1,type='n')
points(Forecast10$Nf[ss,1,1],Forecast10$Nf[ss,3,1],cex=0.5,col="purple",pch=18)
points(Forecast10$Nf[ss,1,2],Forecast10$Nf[ss,3,2],cex=0.5,col="violet",pch=18)
points(Analysis10$Na[ss,1],Analysis10$Na[ss,3],pch=Analysis10$Ca[ss],col=ecoforecastR::col.alpha("#b2df8a",alpha=0.8),cex=0.5)
points(Analysis10$Na[ss,4],Analysis10$Na[ss,6],pch=Analysis10$Ca[ss],col=ecoforecastR::col.alpha("green",alpha=0.8),cex=0.5)
points(No[10,1],No[10,3],cex=3,col=1)
lines(No[10,1]+c(-1,1)*sqrt(R[1,1]),rep(No[10,3],2))
lines(rep(No[10,1],2),No[10,3]+c(-1,1)*sqrt(R[3,3]))
points(No[10,4],No[10,6],cex=3,col=1)
lines(No[10,4]+c(-1,1)*sqrt(R[1,1]),rep(No[10,6],2))
lines(rep(No[10,4],2),No[10,6]+c(-1,1)*sqrt(R[3,3]))

@@

Prior to applying the VSEM to simulate the conifer forest carbon cycle in Oregon the model was calibrated against AGB, leaf C, and NEE data from Ameriflux site US-ME2.  Best fit parameters are given in Supplemental Table \@ref(tab:parameters). The best fit to the observed data was for AGB, which had a $R^2$ of 0.977, a RMSE of 0.089 $kgC/m^2$ (1.7%), and a bias of -1.1% (-0.059 $kgC/m^2$) (Supplemental figure \@ref(fig:calibration) ). Achieving this fit came at the cost of nontrivial systematic biases in both leaf carbon (0.064 $kgC/m^2$, 13%) and NEE (0.528 $kgC/m^2/year$, 69%). The RMSEs for leaf carbon (0.067 $kgC/m^2$, 13.6%) and NEE (0.612 $kgC/m^2/year$, 80%) were similar to the biases, suggesting that most of the model error was systematic, particularly for leaf carbon.

@@

```{r oregon, echo=FALSE, fig.asp=1, fig.cap="Filter divergence and disturbance detection for central Oregon test case. Left column is the default configuration while the right column is a reduced observation error experiment (Exp 2). Top row is the assimilation for an example site. Middle row shows the across-site distribution of divergence metrics and how these vary by disturbance magnitude (purple) and disturbance type (symbols). Bottom row shows how the probability of detecting the disturbance varies with disturbance magnitude"}
## summary figure showing didn't work great at default error rate
m = matrix(1:6,nrow=3,byrow = TRUE)
layout(m)

### Row 1: Example time-series ###
site = 188
exp.name = c("P1Q1R1","P1Q1R4")
exp.pretty.name = c("(a) Example site, Original","(c) Example site, Low Obs Error")
pool.names=c("Leaf","Soil","AGB")
i = 3 ## Stem
time = 1985:2017
for(ex in seq_along(exp.name)){
  load(paste0("OR_TS_",site,"_",exp.name[ex],".RData"))
  plot(time,Yi[,i],
       ylim = range(rbind(CI.Na,Yi[,i],CI.D),na.rm=TRUE),
       type='n',
       ylab=bquote(.(pool.names[i]) ~ (kgC/m^2)),xlab="time",
       main=exp.pretty.name[ex])
  ecoforecastR::ciEnvelope(time,CI.Na[1,],CI.Na[5,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
  lines(time,CI.Na[3,],col="#b2df8a",lwd=2)
  points(time,Yi[,i])
  if(ex == 1) {
    legend("bottomleft",legend=c("Assimilation","Data"),
       lwd=5,col=c("#b2df8a","black"),cex=0.7)
  }
}

@@

### Row 3: Disturbance magnitude vs P(detection) ###
pred.seq=list()
distabs = obs.pre-obs.dist
pred.seq$distabs = seq(min(distabs),max(distabs),length=500)
detect.pretty.name = c("(c) Probability of detection, Original","(f) Probability of Detection, Low Obs Error")
dist.qb.abs <- list()
for(i in 1:2){
  distrate = distrate = elist[[i]]$distrate
  dist.qb.abs[[i]] = glm(distrate ~ distabs,
                    na.action=na.omit,
                    family=quasibinomial("logit")
  )
  plot(distabs,distrate,col=disturb.col.al[obs.class.re],
       xlab="AGB disturbance magnitude",ylab="P(detect)",
       pch=16, main=detect.pretty.name[i],xlim=c(0,17))
  lines(pred.seq$distabs,predict(dist.qb.abs[[i]],newdata=pred.seq,type="response"),col="purple")
  if(i == 1) legend("topleft",legend=disturb.names,
                  col=disturb.col,pch=16,cex=0.7,
                  text.col=c(1,disturb.col[2:5]))
}
```
In total we assimilated 365 example AGB timeseries from our central Oregon study region, with 81 coming from undisturbed sites and 78, 88, 92, and 17 coming from clear cuts, fires, forest pests, and other disturbances, respectively. Initial applications of the multinomial filter exhibited high rates of filter divergence, with a coverage of only `r format(stats[[1]]$div.cov*100,digits=3)`% of the observations falling within the Analysis 95% CI at the end of the timeseries. Filter divergence is a known issue for Kalman Filters, not something unique to our multinomial filter, and results from the model becoming falsely overconfident in itself such that it ignores, and diverges from, the data. Figure \@ref(fig:oregon) a demonstates an example of this for a site that experienced a forest fire in 1999, yet the model continues to believe the site is undisturbed. Across sites the magnitude of filter divergence observed was negatively correlated with the amount of biomass at the end of the timeseries (Figure \@ref(fig:oregon) b, p < `r format(anova(div.reg[[1]])$"Pr(>F)"[1],digits=3,scientific=TRUE)`). In other words, sites with low final biomass (i.e., large previous disturbance) showed a larger divergence and the divergence declined as the final biomass increased. Divergence also varied by disturbance type (p < `r format(anova(div.class[[1]])$"Pr(>F)"[2],digits=3,scientific=TRUE)`), with clear cuts having the lowest divergence for a given disturbance size and forest pests being lower than the fires and "other". Despite the issue of filter divergence there was still a positive relationship between disturbance magnitude and the posterior probability of detecting disturbance (Figure \@ref(fig:oregon) c, p < `r format(summary(dist.qb.abs[[1]])$coefficients[2,4],digits=3,scientific=TRUE)`). However, detection was rarely successful for AGB disturbances that removed less than 7.5 $kgC/m^2$ and did not exceed 50% until disturbances were larger than `r bm=coef(dist.qb.abs[[1]]); format(-bm[1]/bm[2],digits=2)` $kgC/m^2$.

The causes of filter divergence are known to be some combination of model overconfidence (e.g., insufficient process error, Q) and high observation error (R) in the data. As described in the Methods we ran additional computational experiments where we varied Q and R (\@ref(tab:QRexp)) to verify that without filter divergence the multinomial filter was able to capture real-world disturbances and to assess what level of variance inflation was required to do so. Figure \@ref(fig:oregon) d-f show the results for Experiment 2, where we experimented with reducing the observation error, R. In this experiment we see that the data assimilation follows the observations closely, with a much tighter interval estimate, and is able to catch the disturbance event (Figure \@ref(fig:oregon) d). We also see much lower incidence of filter divergence across sites, with a posterior coverage at the final timestep of `r format(stats[[2]]$div.cov*100,digits=3)`%, (Figure \@ref(fig:oregon) e). Divergence is no longer significantly affected by disturbance magnitude (p < `r format(summary(div.class[[2]])$coefficients[2,4],digits=3)`) and the only significant effect of disturbance type was a small reduction in divergence for clearcuts (p < `r format(summary(div.class[[2]])$coefficients[3,4],digits=3)`). Most importantly, detection probabilities were strongly related to disturbance magnitude (Figure \@ref(fig:oregon) f), successfully detecting disturbances that removed more than `r bm=coef(dist.qb.abs[[2]]); format(-bm[1]/bm[2],digits=2)` $kgC/m^2$.

@@

```{r divExp,echo=FALSE,fig.cap="Sensitivity of (a) filter divergence coverage, and (b) disturbance detection threshold to variance inflation of the observation error (x-axis) and process error (symbols)",fig.asp=1 , fig.width=4}
m = matrix(1:2,nrow=2,byrow = TRUE)
layout(m)
## Divergence
div.cov = sapply(stats,"[[","div.cov")
plot(Rmult[Rs],div.cov,col=Qs,pch=Qs,log="x",xlab="Variance Multiplier",ylab="Filter Coverage",main="(a) Filter Coverage")
legend("topright",legend=paste0(Qmult[1:2],"x"),col=1:2,pch=1:2,title="Q")

@@

## Detection probability
for(i in 3:6){
  distrate = distrate = elist[[i]]$distrate
  dist.qb.abs[[i]] = glm(distrate ~ distabs,
                    na.action=na.omit,
                    family=quasibinomial("logit")
  )
}
thresh = sapply(dist.qb.abs,function(x){bm=coef(x); -bm[1]/bm[2]})
plot(Rmult[Rs],thresh,col=Qs,pch=Qs,log="x",xlab="Variance Multiplier",ylab=expression("Detection Threshold" ~ (kgC/m^2)),main="(b) Disturbance Detection Threshold")
legend("topleft",legend=paste0(Qmult[1:2],"x"),col=1:2,pch=1:2)
```
